cardinalR: Generating interesting high-dimensional data structures

### 2025-09-16 paper-cardinalR.Rmd
Cube section rework

@@

#### `gen_gridcube()`

The `gen_gridcube(n, p)` constructs a regular grid using approximately $n$ points by evenly partitioning each of the $p$ dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cube) a"} else { "Figure \\@ref(fig:cube-proj) a"})`). Specifically, the number of grid points along each axis is determined by `gen_nproduct(n, p)`, which calculates the nearest balanced integer factors whose product is close to $n$. The full grid is generated via `expand_grid()`, giving Cartesian combinations of these evenly spaced positions.

Each variable $X_j$ represents integer grid coordinates (e.g., 1, 2, 3, ...) in dimension $j$: $X_j \in {1, \dots, k_j},\text{ where }\prod_{j=1}^p k_j \approx n$. This results in a hypercube-shaped lattice in $p\text{-}D$.

@@

#### `gen_unifcube()`

The `gen_unifcube(n, p)` function generates $n$ points uniformly distributed within a $3\text{-}D$ cube centered at the origin, optionally embedded in $p\text{-}D$ with additional noise dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cube) b"} else { "Figure \\@ref(fig:cube-proj) b"})`).

Points are sampled independently from $X_1, X_2, X_3 \sim U(-0.5, 0.5)$ of a cube of side length $1$. This results in a uniform cloud of points filling the unit cube in $3\text{-}D$ space. For $p > 3$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $3\text{-}D$ structure into $p\text{-}D$, where $j = 4, \dots, p$.

@@

#### `gen_cubehole()`

The `gen_cubehole(n, p)` function generates a dataset of $n$ points sampled uniformly inside a cube, then removes points that fall within a central spherical hole, resulting in a hollow cube structure in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cube) c"} else { "Figure \\@ref(fig:cube-proj) c"})`).

The initial points are drawn from a uniform cube centered at the origin: $X_1, X_2, X_3 \sim U(-0.5, 0.5)$. The center of the cube is $(0, 0, 0)$. A spherical hole of radius $0.5$ is carved out by filtering out points whose Euclidean distance from the origin is less than or equal to $0.5$: remove if $\sqrt{X_1^2 + X_2^2 + X_3^2} \leq 0.5$. This results in a shell-like point cloud that fills the cube but avoids the spherical center. For $p > 3$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $3\text{-}D$ structure into $p\text{-}D$, where $j = 4, \dots, p$.

>>>>

The first is the regular grid of points of $n$ points in $p$ dimensions. This is generated using `gen_gridcube(n, p)`. The number of grid points along each axis is determined by finding the nearest integer factors whose product is close to $n$. In the base cube function, the cube is represented in integer grid coordinates (e.g. 1, 2, 3, ...) along each ais. This results in a $p\text{-}D$ hyercube-shaped lattice structure.

An extension to the regular grid of points is to consider the points being uniformly distributed along each axis, as opposed to evenly spaced. The function `gen_unifcube(n, p)` is identical to the regular grid of points, except instead of points being placed in integer grid coordinates, they are placed at a uniformly distributed point inside the $p\text{-}D$ cube (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cube) b"} else { "Figure \\@ref(fig:cube-proj) b"})`).

Finally, we consider a cube that has a central spherical hole. This is generated using the `gen_cubehole(n, p)` function. The cube is generated as per the uniformly distributed cube, but points inside sphere of radius $0.5$ are removed, resulting in a hollow cube structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cube) c"} else { "Figure \\@ref(fig:cube-proj) c"})`).

@@

@@

@@
